---
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
bibliography: references.bib
csl: associacao-brasileira-de-normas-tecnicas.csl
urlcolor: black
linkcolor: black
fontsize: 12pt
geometry: margin = 1.2in
header-includes:
- \usepackage{placeins}
- \usepackage{fancyhdr}
- \usepackage{setspace}
- \usepackage{chngcntr}
- \usepackage{microtype}
- \usepackage{dcolumn}
- \onehalfspacing
- \counterwithin{figure}{section}
- \counterwithin{table}{section}
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path = '00_dissertation/figures/',
                      echo = FALSE, warning = FALSE, message = FALSE)
```

\pagenumbering{gobble}

```{r child = '00a_title_page.Rmd'}
```

\newpage
\pagestyle{fancy}

\fancyhead[LE,RO]{}
\fancyhead[LO,RE]{}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

\pagenumbering{roman}

\fancyhead[CO,CE]{Abstract}
\section*{Abstract}
\addcontentsline{toc}{section}{Abstract}

```{r child = '00b_abstract.Rmd'}
```

\newpage

\fancyhead[CO,CE]{Table of Contents}
\setcounter{tocdepth}{2}
\tableofcontents

\newpage

\fancyhead[CO,CE]{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures

\newpage
\fancyhead[CO,CE]{List of Tables}
\addcontentsline{toc}{section}{List of Tables}
\listoftables

\newpage
\pagenumbering{arabic}
\fancyhead[CO,CE]{Introduction}

```{r child = '00c_intro.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Chapter 1 - The premium of territorialization strategies}

```{r child = '01_premium.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Chapter 2 - The use(s) of individual budget amendments}

```{r child = '02_iba.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Chapter 3 - The microrelations of nonprofit-public partnerships}

```{r child = '03_nppp.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{Conclusion}

```{r child = '04_conclusion.Rmd'}
```

\FloatBarrier
\newpage
\fancyhead[CO,CE]{References}

# References